---
title: "Datasets"
author: "Kevin Middleton"
date: "1/12/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cowplot)
```

See:

- https://twitter.com/CoolDatasets
- https://github.com/caesar0301/awesome-public-datasets
- http://notawfulandboring.blogspot.com/2016/12/kevin-mcintyres-open-stats-lab.html
- https://coast.noaa.gov/dataregistry/search/collection
- https://www.nodc.noaa.gov/cgi-bin/OAS/prd/accession/download/115356

## `Lahman`: Sean Lahman's Baseball Database

- See: http://varianceexplained.org/r/simulation-bayes-baseball/
- Height, weight, batting, pitching for 18000+ baseball players.

```{r}
library(tidyverse)
library(Lahman)
career <- Batting %>%
  filter(AB > 0) %>%
  anti_join(Pitching, by = "playerID") %>%
  group_by(playerID) %>%
  summarize(H = sum(H), AB = sum(AB)) %>%
  mutate(average = H / AB)

career <- Master %>%
  tbl_df() %>%
  select(playerID, nameFirst, nameLast) %>%
  unite(name, nameFirst, nameLast, sep = " ") %>%
  inner_join(career, by = "playerID") %>%
  select(-playerID)
career
```

```{r}
Batting %>%
  filter(AB > 500) %>%
  anti_join(Pitching, by = "playerID") %>%
  group_by(playerID, yearID) %>%
  summarize(H = sum(H), AB = sum(AB)) %>%
  mutate(average = H / AB) %>%
  ungroup() %>% 
  select(yearID, average) %>% 
  group_by(yearID) %>% 
  summarize(max_avg = max(average)) %>% 
  ggplot(aes(yearID, max_avg)) +
  geom_line() +
  labs(y = "Highest Batting Average")
```

## STDs nationally ranked by state by year

- 2000-2015

```{r}
library(tidyverse)
STD <- read_csv("data/STDs_Nationally_Ranked_By_State.csv")
glimpse(STD)
```

```{r}
STD %>% ggplot(aes(x = Year, y = Chlamydia_Rate, color = State)) +
  geom_line() +
  labs(y = "Rate of Chlamydia")
```

## Birth rates

- https://www.datazar.com/project/p8801c779-35d2-4d2b-add2-087e11a67298
- By year and age group, 1940-2013

```{r}
library(tidyverse)
Births <- read_csv("data/birth_rates.csv")
Births %>% 
  ggplot(aes(Year, Birth_Rate, color = Age_Group)) +
  geom_line() +
  labs(y = "Birth Rate")
```

## Probability of death

- Males, 1900 to present

```{r}
P_death <- read_csv("data/Death_Probability_of_Males_Since_1900.csv")

P_death %>% 
  filter(Year == 1900) %>% 
  gather(key, value, -Year) %>% 
  mutate(key = as.numeric(key)) %>% 
  ggplot(aes(y = value, x = key)) +
  geom_bar(stat = "identity") +
  labs(y = "Probability of death", x = "Age")

P_death %>% 
  select(Year, `0`) %>% 
  ggplot(aes(Year, `0`)) + 
  geom_line() +
  labs(y = "Probability of death before age 1")
```

## NASA fireball and bolide incidents

```{r}
library(tidyverse)
library(stringr)
library(ggmap)

fireball <- read_csv("data/NASA_Fireball_And_Bolide_Reports.csv")
head(fireball)

sanitize_latitude <- function(x) {
  NS <- str_sub(x, start = -1)
  lat <- as.numeric(str_replace(x, "[NS]", ""))
  lat <- if_else(NS == "S", -lat, lat)
  return(lat)
}

sanitize_longitude <- function(x) {
  EW <- str_sub(x, start = -1)
  long <- as.numeric(str_replace(x, "[EW]", ""))
  long <- if_else(EW == "W", -long, long)
  return(long)
}

fireball <- fireball %>% 
  mutate(lat = sanitize_latitude(`Latitude (Deg)`),
         long = sanitize_longitude(`Longitude (Deg)`))

map <- get_map(location = c(0, 0), zoom = 1)

ggmap(map) + 
  geom_point(data = fireball,
             aes(long, lat, size = `Total Radiated Energy (J)`)) +
  labs(x = "Longitude", y = "Latitude")
```

## Global landslide catalog export

```{r}
library(tidyverse)
library(ggmap)
landslide <- read_csv("data/Global_Landslide_Catalog_Export.csv") %>% 
  select(latitude, longitude, country)
glimpse(landslide)
```

```{r}
map <- get_map(location = c(-84, 10), zoom = 3)

ggmap(map) + 
  geom_point(data = landslide,
             aes(y = latitude, x = longitude)) +
  labs(x = "Longitude", y = "Latitude")
```

## Los Angeles water use

```{r}
library(tidyverse)
library(stringr)
LA <- read_csv("data/Los_Angeles_Water_Use_Average_By_Zipcode.csv") %>%
  mutate(zip_code = str_split(`Location 1`, "\n", simplify = TRUE)[, 1],
         lat_lon = str_split(`Location 1`, "\n", simplify = TRUE)[, 2])
LA %>% 
  select(-lat_lon, -`Location 1`) %>% 
  gather(key, value, -zip_code) %>% 
  ggplot(aes(x = key, y = value, group = zip_code, color = zip_code)) +
  geom_line() +
  theme(legend.position = "none")
```

## Meteorite landings

```{r}
meteor <- read_csv("data/Meteorite_Landings.csv") %>% 
  select(reclat, reclong, `mass (g)`) %>% 
  drop_na()

map <- get_map(location = c(0, 0), zoom = 1)

ggmap(map) +
  ggplot(data = meteor, aes(y = reclat, x = reclong)) +
  geom_point()

```

## NYC SAT

```{r}
library(tidyverse)
SAT <- read_csv("data/New_York_City_SAT_Results.csv",
                na = "s") %>% 
  drop_na()

SAT %>% 
  ggplot(aes(`SAT Critical Reading Avg. Score`,
             `SAT Writing Avg. Score`)) +
  geom_point()

SAT %>% 
  ggplot(aes(`Num of SAT Test Takers`,
             `SAT Writing Avg. Score`)) +
  geom_point()

SAT %>% 
  ggplot(aes(`Num of SAT Test Takers`,
             `SAT Math Avg. Score`)) +
  geom_point()

SAT %>% 
  select(-DBN, -`SCHOOL NAME`) %>% 
  gather(key, value, -`Num of SAT Test Takers`) %>% 
  ggplot(aes(`Num of SAT Test Takers`, value, color = key)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE)
```

```{r}
with(SAT, cor.test(`SAT Critical Reading Avg. Score`,
                   `SAT Writing Avg. Score`))
with(SAT, cor.test(`SAT Critical Reading Avg. Score`,
                   `SAT Math Avg. Score`))
```

## Fuel Economy

```{r}
library("tidyverse")
M <- read_csv("data/Vehicle_Fuel_Economy.csv")
glimpse(M)
M %>% 
  ggplot(aes(highway08)) +
  geom_histogram()

M %>% 
  filter(highway08 > 75) %>% 
  select(make, model, year, highway08) %>% 
  arrange(year)
```

## fivethirtyeight

https://cran.r-project.org/web/packages/fivethirtyeight/vignettes/fivethirtyeight.html

```{r eval=FALSE}
install_github("rudeboybert/fivethirtyeight")
```

```{r}
library("tidyverse")
library("fivethirtyeight")
# http://fivethirtyeight.com/features/you-cant-trust-what-you-read-about-nutrition/
glimpse(nutrition_pvalues)
nutrition_pvalues %>% 
  ggplot(aes(p_values)) +
  geom_line(stat = "density")
```

