---
title: "Data Loading"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King and Kevin Middleton'
date: 'Last updated: `r Sys.Date()`'
output:
  ioslides_presentation:
    fig_width: 7
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
```

## Notes

## Readings

## Reading delimited files into R

- csv: `read.csv()`
- tab or otherwise delimited: `read.table()`
- [readr](https://cran.r-project.org/web/packages/readr/index.html), loaded with `tidyverse`
    - `read_delim()`
    - `read_csv()`
    - Like their base R counterparts but with more logical default arguments & behavior

## Beware of `stringsAsFactors` "feature"

- Base R file readers (`read.csv()`, `read.table()`, etc.) automatically convert strings into factors (`stringsAsFactors = TRUE`).
- *Probably* what you want, but it can be a giant pain if you don't.
    - You likely don't want your "Notes" column as a factor

## Beware of `stringsAsFactors` "feature"

Options:

- Pass `stringsAsFactors = FALSE` as a parameter.
- Use `readr` functions `read_delim()` and `read_csv()`.

Fun reading:

- [stringsAsFactors = sigh](http://notstatschat.tumblr.com/post/124987394001/stringsasfactors-sigh)
- [stringsAsFactors: An unauthorized biography](http://simplystatistics.org/2015/07/24/stringsasfactors-an-unauthorized-biography/)

## Reading Excel files into R

Preferred package: [readxl](https://cran.r-project.org/web/packages/readxl/index.html)

- *Not* automatically loaded with `tidyverse`

```{r eval=FALSE}
library("readxl")

M <- read_excel("MyFile.xlsx")
```

## Load some data

```{r message=FALSE, warning=FALSE}
M <- read_excel("../data/Mouse_Weaning_Data.xlsx",
                na = "NA")
```

Read file and assign to an object `M`.

- Look for it in the `data` directory, which is one level up from the working directory
- Explicitly say that "NA" is `NA`

Use `<-` for assign.

- Don't use `=` (even though you can).
- "Assign the object on the right to the object on the left."

## Use `str()` to get information

```{r}
str(M)
```

## Selecting and filtering

- Selecting columns
- Filtering a subset of rows
- Aggregating summary statistics

## Selecting columns: Base R

```{r eval=FALSE}
M$WnMass
```

`$` returns a single column from a `data.frame`. You'll use this form most often day to day.

## Selecting columns: Base R

```{r eval=FALSE}
M[ , "WnMass"]  # Only the column WnMass
M[ , 1:3]       # Columns 1-3, all rows
```

`[rows, columns]` returns rows and columns from a `data.frame`.

If either is empty, all are returned.

- All rows are returned here. The space before `,` is empty.

## Filtering rows: Base R

```{r eval=FALSE}
M[1, ]     # Row 1, all columns
M[1:5, ]   # Rows 1-5, all columns
```

- The space after `,` is empty, so all columns are returned.

## Filtering rows: Base R

Use `head()` to get a peek at the first few rows. A second argument to `head()` specifies the number of rows to show.

```{r}
head(M[, c("MouseID", "Sex")])
```

## Select and filter simultaneously: Base R

`subset()` or `[ , ]`

```{r}
subset(M, WnMass < 6)
```

## Select and filter simultaneously: Base R

`subset()` or `[ , ]`

```{r eval=FALSE}
subset(M, Sex == 1 & DamID == 1932, select = "WnMass")
```

Use `&` for "and", and `|` for "or."

## Making new columns

- Calculate new variables
    - Transformation, etc.
- Combine existing variables into a new variable
    - e.g., year, months, day columns into a date
- Splitting existing variables into >1 new variables

## Calculate new variables: Base R

Data transformations:

```{r eval=FALSE}
M$log10_WnMass <- log10(M$WnMass)
M$sqrt_WnMass <- sqrt(M$WnMass)
```

_Note_:

- `log()` performs natural log.
- `log10()` performs base-10 log.

## Aggregating: Base R

- Mean of every column that is not `DamID`. `head()` gives the first few rows.
- Replace `mean` with `sd` to get standard deviation.

```{r warning=FALSE}
M_agr <- aggregate(M, by = list(M$DamID), FUN = "mean")
head(M_agr)
```

## New tools: Selecting, filtering, aggregating

- [tidyr](https://cran.r-project.org/web/packages/tidyr/index.html)
- [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html)
- Both load with `tidyverse`

Tutorials:

- http://seananderson.ca/2014/09/13/dplyr-intro.html
- https://rpubs.com/bradleyboehmke/data_wrangling

```{r}
library("tidyverse")
```

## tidyr

tidyr is most useful for moving data between "wide" and "long" formats:

- `gather()`: wide to long
- `spread()`: long to wide

Utility functions for existing variables:

- `separate()`: splitting variables (1 column to >1)
- `unite()`: joining variables (>1 columns to 1)

## Combine existing variables: tidyr

```{r}
M <- M %>% unite(col = "YearMo", WnYear, WnMonth,
                 sep = "_", remove = FALSE)
M %>% select(WnYear, WnMonth, YearMo) %>% head()
```

`remove = FALSE` means that the pre-existing variables are not removed.

## Main functions in dplyr

- `mutate()`: create new columns
- `glimpse()`: like `str()`
- `select()`: like `[ , ]` for columns
- `filter()`: like `subset()` or `[ , ]` with conditions
- `summarize()` with `group_by()`: like `aggregate()`
- `arrange()`: sorting
- `join()`: merging `data.frames`

Functions can (should) be chained together with `%>%`

- "Pipe" the result of the lefthand expression to the righthand expression.

## Calculate new variables: dplyr

```{r eval=FALSE}
M <- M %>%
  mutate(log10DamMass = log10(DamMass))
```

- Assigning back to itself just adds the column `log10DamMass` to the end of the `data.frame`.

## `mutate()` or not?

`mutate()` lets you add many new variables at once.

```{r eval=FALSE}
M$log10_WnMass <- log10(M$WnMass)
M$log10_DamMass <- log10(M$DamMass)
```

vs.

```{r}
M <- M %>%
  mutate(log10WnMass = log10(WnMass),
         log10DamMass = log10(DamMass))
```

## `glimpse()` is like `str()`

```{r}
glimpse(M)
```

## Selecting columns: dplyr

```{r}
M %>% select(Sex, WnMass, DamID)
```

## Chaining operations

```{r}
M %>% select(starts_with("wn")) %>% glimpse()
```

For more information and examples: `?select`

`starts_with()`, `ends_with()`, `contains()`, `matches()`, `num_range()`, `one_of()`, `everything()`

## Benefits of chaining operations

1. Cleaner code, reads like a sentence
    - Be wary of long, opaque, chains (comment, comment, comment) 
2. (Often) avoid saving intermediate objects.
    - Select, filter, and plot in one step 

## Filtering: dplyr

```{r}
M %>% filter((WnMass < 6 | WnMass > 19) & Sex == 0)
```

## Sorting: dplyr

```{r}
M %>% filter((WnMass < 6 | WnMass > 19) & Sex == 0) %>%
  arrange(WnMass)
```

## Selecting and filtering

```{r}
M %>% filter((WnMass < 6 | WnMass > 19) & Sex == 0) %>% 
  select(MouseID, DamID, WnMass)
```
## Quiz 02-2

On Canvas, complete quiz 02-2.

Move on to lecture 02-3.
