---
title: "Data Visualization"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King and Kevin Middleton'
date: 'Last updated: `r Sys.Date()`'
output:
  xaringan::moon_reader:
    seal: yes
    nature:
      highlightStyle: github
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE}
```

---

# Notes

---

# Readings

---

## Plotting: Get to know ggplot {.smaller}

- ggplot documentation reference: http://docs.ggplot2.org/current/
- R Graphics Cookbook: http://www.cookbook-r.com/Graphs/
- Quick-R: http://www.statmethods.net/advgraphs/ggplot2.html
- Quick Introduction to ggplot: http://blog.echen.me/2012/01/17/quick-introduction-to-ggplot2/
- ggplot Cheatsheet: http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/
- Tutorial: http://www.ceb-institute.org/bbs/wp-content/uploads/2011/09/handout_ggplot2.pdf
- Advanced: http://rpubs.com/hadley/97970
- https://github.com/jennybc/ggplot2-tutorial
- https://speakerdeck.com/jennybc/ggplot2-tutorial
- http://inundata.org/2013/04/10/a-quick-introduction-to-ggplot2/

## Other R plotting systems

Base R graphics (e.g., `plot()`)

- Ugly
- Very verbose

Lattice (`library(lattice)`)

- Largely overlappying functionality with ggplot
- Much less friendly interface (really)

## Basics of ggplot

Anderson, E. 1935. The irises of the Gaspe Peninsula. *Bulletin of the American Iris Society* 59: 2–5.

```{r}
glimpse(iris)
```


## Basics of ggplot

```{r echo=TRUE, eval=FALSE}
library("ggplot2")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) + 
  geom_point() 
```

- Load the `ggplot` package
- `ggplot()` sets up the plot using `iris` data.frame.
- `aes()` sets `x` and `y` variables
- `geom_point()` adds `(x, y)` points corresponding to `Petal.Length` and `Petal.Width`

## Logic of ggplot

- Building plots up by layers
    - Start simple
- Layers can be:
    - data
    - labels
    - colors
    - regression lines, etc.
- Change the theme (shading, colors, line widths, etc.)
- You might see examples with `qplot()`. Ignore them. Learn ggplot the hard way. Shortcuts will limit you in the long run.

## Logic of ggplot

```{r eval=FALSE}
library("ggplot2")
library("cowplot")  # Easy on the eyes ggplot theme
ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point() 
```

## Basics of ggplot

```{r echo=FALSE}
library("ggplot2")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point() 
```

## What would setting only 1 variable do?

```{r eval=FALSE}
ggplot(iris, aes(x = Petal.Length)) +
```

## What would setting only 1 variable do?

```{r warning=FALSE, message=FALSE}
ggplot(iris, aes(x = Petal.Length)) + 
  geom_histogram() 
```

## Building plots

```{r echo=TRUE, eval=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point()
```

- `color = Species` colors points by the _Species_ column (a categorical variable)

## Building plots

```{r echo=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point()
```

## Building plots

```{r eval=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 shape = Species)) + 
  geom_point()
```

- `shape = Species` codes _Species_ by the point shape

## Building plots

```{r echo=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 shape = Species)) + 
  geom_point()
```

## Building plots

```{r echo=TRUE, eval=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 color = Species)) + 
  geom_point(size = 3)
```

- Change point size

## Building plots

```{r echo=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point(size = 3)
```

## Building plots

```{r echo=TRUE, eval=FALSE}
library("wesanderson")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 color = Species)) + 
  geom_point(size = 3) + 
  scale_color_manual(values = wes_palette("Moonrise2"))
```

- Add an interesting color palette

## Building plots

```{r echo=FALSE}
library("wesanderson")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point(size = 3) + 
  scale_color_manual(values = wes_palette("Moonrise2"))
```

## Building plots

```{r echo=TRUE, eval=FALSE}
library("wesanderson")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 color = Species)) + 
  geom_point(size = 3) + 
  scale_color_manual(values = wes_palette("Moonrise2")) +
  xlab("Petal Length (cm)") +
  ylab("Petal Width (cm)") +
  ggtitle("Irises")
```

- Add some better axis labels

## Building plots

```{r echo=FALSE}
library("wesanderson")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point(size = 3) + 
  scale_color_manual(values = wes_palette("Moonrise2")) +
  xlab("Petal Length (cm)") +
  ylab("Petal Width (cm)") +
  ggtitle("Irises")
```

## Adding lines

```{r echo=TRUE, eval=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

- `geom_line()` add a "smooth" line (but in this case linear [`lm`])

## Adding lines

```{r echo=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

## Adding lines

What if you add a color aesthetic?

```{r echo=TRUE, eval=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

## Adding lines

```{r echo=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

## Log-log axes

```{r eval=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10()
```

## Log-log axes

```{r echo=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10()
```

## Facetting

```{r eval=FALSE}
ggplot(iris, aes(x = Petal.Length)) + 
  geom_histogram() +
  facet_grid(Species ~ .)
```

- Plots separate histograms for the 3 species in 3 rows
- `facet_grid(. ~ Species)` makes 3 columns

## Facetting

```{r echo=FALSE, message=FALSE}
ggplot(iris, aes(x = Petal.Length)) + 
  geom_histogram() +
  facet_grid(Species ~ .)
```

## `_geoms` we will use

- `geom_point()`
- `geom_smooth()`
- `geom_line()`
- `geom_histogram()`
- `geom_boxplot()`
- `geom_dotplot()`
- More at http://docs.ggplot2.org/current/
