---
title: "Data Visualization"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King, Kevin Middleton and Lauren Sullivan'
output:
  ioslides_presentation:
    fig_width: 7
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(cowplot)
theme_set(theme_cowplot())
```

## R plotting systems

1. Base R graphics (e.g., `plot()`, `hist()`)
    - Easy to do easy things (bivariate scatterplot, histogram)
    - Difficult to do complex plots
2. ggplot
    - Harder to learn, easier to make complex plots
    - Much better graphics and logic


## Basics of ggplot

ggplot graphics

- [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html), loaded with `tidyverse`

- Anderson, E. 1935. The irises of the Gaspe Peninsula. *Bulletin of the American Iris Society* 59: 2–5.

```{r}
glimpse(iris)
```

## Basics of ggplot

```{r echo=TRUE, eval=FALSE}
library("tidyverse")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) + 
  geom_point() 
```

- `ggplot()` sets up the plot using `iris` data.frame.
- `aes()` sets `x` and `y` variables
- `geom_point()` adds `(x, y)` points corresponding to `Petal.Length` and `Petal.Width`

## Logic of ggplot

- Building plots up by layers
    - Start simple
- Layers can be:
    - data
    - labels
    - colors
    - regression lines, etc.
- Change the theme (shading, colors, line widths, etc.)
- You might see examples with `qplot()`. Ignore them. Learn ggplot the hard way. Shortcuts will limit you in the long run.

## Logic of ggplot

```{r eval=FALSE}
library("tidyverse")
library("cowplot")  # Our preferred ggplot theme
ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point() 
```

## Basics of ggplot

```{r echo=FALSE}
library("ggplot2")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point() 
```

## What would setting only 1 variable do?

```{r eval=FALSE}
ggplot(iris, aes(x = Petal.Length)) +
```

## What would setting only 1 variable do?

```{r warning=FALSE, message=FALSE, fig.height=3.5}
ggplot(iris, aes(x = Petal.Length)) + 
  geom_histogram() 
```

## Building plots

```{r pt_color, echo=TRUE, eval=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point()
```

- `color = Species` colors points by the _Species_ column (a factor)

## Building plots

```{r pt_color, echo=FALSE}
```

## Building plots

```{r pt_shape, eval=FALSE}
ggplot(iris, aes(x = Petal.Length,
                 y = Petal.Width,
                 shape = Species)) + 
  geom_point()
```

- `shape = Species` encodes _Species_ by the point shape

## Building plots

```{r pt_shape, echo=FALSE}
```

## Building plots

```{r pt_size, echo=TRUE, eval=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 color = Species)) + 
  geom_point(size = 3)
```

- Change point size

## Building plots

```{r pt_size, echo=FALSE}
```

## Building plots

```{r plot_colors, echo=TRUE, eval=FALSE}
library("wesanderson")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 color = Species)) + 
  geom_point(size = 3) + 
  scale_color_manual(values = wes_palette("Moonrise2"))
```

- Add an interesting color palette

## Building plots

```{r plot_colors, echo=FALSE}
```


## Building plots

```{r plot_colors_colorblind, echo=TRUE, eval=FALSE}
library("viridis")
ggplot(iris, aes(x = Petal.Length, y = Petal.Width,
                 color = Species)) + 
  geom_point(size = 3) + 
  scale_color_viridis(discrete = TRUE, option = "viridis")
```

- `library(viridis)` works for colorblind and grayscale friendly images.


## Building plots
```{r plot_colors_colorblind, echo=FALSE, message = FALSE}
```


## Building plots

```{r better_axis_labels, echo=TRUE, eval=FALSE}
library("wesanderson")
ggplot(iris, aes(x = Petal.Length,
                 y = Petal.Width,
                 color = Species)) + 
  geom_point(size = 3) + 
  scale_color_manual(values = wes_palette("Moonrise2")) +
  labs(x = "Petal Length (cm)",
       y = "Petal Width (cm)",
       title = "Irises")
```

- Add some better axis labels

## Building plots

```{r better_axis_labels, echo=FALSE}
```

## Adding lines

```{r add_lm, echo=TRUE, eval=FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

- `geom_line()` add a "smooth" line (but in this case linear [`lm`])

## Adding lines

```{r add_lm, echo=FALSE}
```

## Adding lines

What if you add a color aesthetic?

```{r lm_color, echo=TRUE, eval=FALSE}
ggplot(iris, aes(x = Petal.Length,
                 y = Petal.Width,
                 color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

## Adding lines

```{r lm_color, echo=FALSE}
```

## Log-log axes

```{r loglog, eval=FALSE}
ggplot(iris, aes(x = Petal.Length, 
                 y = Petal.Width,
                 color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10()
```

## Log-log axes

```{r loglog, echo=FALSE}
```

## Facetting

```{r facets, eval=FALSE}
ggplot(iris, aes(x = Petal.Length)) + 
  geom_histogram() +
  facet_grid(Species ~ .)
```

- Plots separate histograms for the 3 species in 3 rows
- `facet_grid(. ~ Species)` makes 3 columns

## Facetting

```{r facets, echo=FALSE, message=FALSE}
```

## `_geoms` we will use

- `geom_point()`
- `geom_smooth()`
- `geom_line()`
- `geom_histogram()`
- `geom_density()`
- `geom_boxplot()`
- `geom_dotplot()`


