---
title: "Data Cleaning"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King and Kevin Middleton'
date: 'Last updated: `r Sys.Date()`'
output:
  ioslides_presentation:
    fig_width: 7
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(cowplot)
```

## Notes

Standard setup chunk:

```{r eval=FALSE}
library(tidyverse)  # Load dplyr, ggplot2, readr, etc.
library(readxl)     # Read Excel files
library(cowplot)    # Prettier ggplot defaults
```

## Readings

## Load data

```{r message=FALSE, warning=FALSE}
M <- read_excel("../data/Mouse_Weaning_Data.xlsx",
                na = "NA")
str(M)
```

## Aggregating: dplyr

Start by grouping data by some categorical variable:

```{r}
M %>% group_by(Sex)
```

## Aggregating: dplyr

Pass the grouped data to `summarize()`. Return the mean and standard deviation.

```{r}
M %>% group_by(Sex) %>% summarize(mean(WnMass), sd(WnMass))
```

Column names `mean(WnMass)` and `sd(WnMass)` are not ideal.

## Aggregating: dplyr

```{r}
M %>% group_by(Sex) %>%
  summarize(mass_mean = mean(WnMass),
            mass_sd = sd(WnMass))
```

What's going on with `Sex` = `NA`?

## Counting `NA`s

```{r}
sum(is.na(M$Sex))
```

- `is.na()` returns `TRUE` if the value is `NA`
- Summing them converts `TRUE` to 1

We should probably drop those rows or go back to the original data sheets.

## Filtering out unwanted rows {.smaller}

```{r}
M_no_NAs <- M %>% filter(!is.na(Sex))
```

`!is.na()` returns `TRUE` for the values of `Sex` _not_ `NA`.

```{r}
sum(is.na(M_no_NAs$Sex))
```

```{r}
M_no_NAs %>% group_by(Sex) %>%
  summarize(mass_mean = mean(WnMass),
            mass_sd = sd(WnMass))
```

## Filtering out unwanted rows {.smaller}

```{r}
M_no_NAs <- M %>% drop_na(Sex)
```

```{r}
sum(is.na(M_no_NAs$Sex))
```

```{r}
M_no_NAs %>% group_by(Sex) %>%
  summarize(mass_mean = mean(WnMass),
            mass_sd = sd(WnMass))
```

## Visualizing data for checking

```{r echo=FALSE, message=FALSE, warning=FALSE}
M_no_NAs <- read_excel("../data/Mouse_Weaning_Data_Bad.xlsx",
                na = "NA") %>%
  drop_na(Sex)
```

```{r message=FALSE, warning=FALSE, fig.height=3.5}
ggplot(M_no_NAs, aes(x = WnMass)) +
  geom_histogram() +
  facet_grid(Sex ~ .)
```

## Finding suspect observations

```{r message=FALSE, warning=FALSE}
M_no_NAs %>% 
  select(MouseID, DamID, WnMass) %>% 
  arrange(WnMass)
```

## Finding suspect observations

```{r}
M_no_NAs %>% 
  select(MouseID, DamID, WnMass) %>% 
  arrange(desc(WnMass))
```

## What to do?

1. Are the observations really incorrect?
    - Malfunction
2. Was there a data entry error?
    - Check the original data sheets
    
## Filter out

```{r}
M_filt <- M_no_NAs %>% 
  filter(WnMass > 2 & WnMass < 100)

max(M_filt$WnMass)
min(M_filt$WnMass)
```

## Replace values

```{r}
M_no_NAs$WnMass[M_no_NAs$MouseID == 207] <- 14.09

M_no_NAs$WnMass[M_no_NAs$MouseID == 327] <- 13.96

M_no_NAs$WnMass[M_no_NAs$MouseID == -9 &
                  M_no_NAs$WnMass == 0.1636] <- 16.36

M_no_NAs$WnMass[M_no_NAs$MouseID == -9 &
                  M_no_NAs$WnMass == 105.20] <- 15.20
```

## Finding suspect observations

```{r}
M_no_NAs %>% 
  select(MouseID, DamID, WnMass) %>% 
  arrange(WnMass)
```

## Finding suspect observations

```{r}
M_no_NAs %>% 
  select(MouseID, DamID, WnMass) %>% 
  arrange(desc(WnMass))
```

## Visualizing data for checking

```{r echo=FALSE, message=FALSE}
ggplot(M_no_NAs, aes(x = WnMass)) +
  geom_histogram() +
  facet_grid(Sex ~ .)
```

## Visualizing data for checking

```{r echo=TRUE, message=FALSE, fig.height=3}
M_no_NAs %>% 
  filter(MouseID != -9) %>% 
  ggplot(aes(y = WnMass, x = factor(Sex))) +
  geom_boxplot()
```

## Data integrity

- Careful checking is the first step in your analysis.
- Treat this as an imperative.

## Quiz 02-4

There is no quiz for this lecture.

Problem Set 2 will be posted on Wednesday.

See you Friday.
