---
title: "Applications of Inference Frameworks: More Basics"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King, Kevin Middleton, and Lauren Sullivan'
output:
  ioslides_presentation:
    fig_width: 8
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())

SticklebackPlates <- read_csv("../data/Stickleback_Plates.csv")

```

## Boxplots

Different variations, but the most standard recipe is:

1. Median = center line
2. Box = IQR (25th to 75th percentiles)
3. Whiskers at 1.5 X 25th and 75th percentiles
4. Dots for anything outside 1.5 X 25th and 75th percentiles

## Boxplots

```{r echo=TRUE, fig.height = 3.5}

ggplot(SticklebackPlates, aes(x = genotype, y = plates)) +
  geom_boxplot()
```

## Raw data as an alternative to boxplots

When the number of observations is small, just show the raw data as points.

```{r eval=FALSE}
ggplot(SticklebackPlates, aes(x = genotype, y = plates)) +
  geom_point()
```

`x = genotype` maps the x position to the categorical variable genotype. `geom_point()` adds the points.

## Raw data as an alternative to boxplots

```{r echo=FALSE}
ggplot(SticklebackPlates, aes(x = genotype, y = plates)) +
  geom_point()
```

## Raw data as an alternative to boxplots

Add a little jitter and transparency on the x axis to keep from overplotting.

```{r eval=FALSE}
ggplot(SticklebackPlates, aes(x = genotype, y = plates)) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.5)
```

## Raw data as an alternative to boxplots

Add a little jitter and transparency on the x axis to keep from overplotting.

```{r echo=FALSE}
ggplot(SticklebackPlates, aes(x = genotype, y = plates)) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.5)
```

## Boxplot and raw data

```{r eval=FALSE}

ggplot(SticklebackPlates, aes(x = genotype, y = plates)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.5)
  
```

## Boxplot and raw data

```{r echo=FALSE}

ggplot(SticklebackPlates, aes(x = genotype, y = plates)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.5)
  
```

- Show them how to add color? or transparency to boxplot? Not sure its necessary
- boxplot is in front now - would block pts without transparency

## Boxplot and raw data

```{r eval=FALSE}

ggplot(SticklebackPlates, aes(x = genotype, y = plates, fill = genotype)) +
  geom_point(position = position_jitter(width = 0.3), alpha = 0.5) + 
  geom_boxplot(alpha = 1/3) 
  
```
