---
title: "Unit 4: In Class Exercises"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King, Kevin Middleton, and Lauren Sullivan'
output:
  ioslides_presentation:
    fig_width: 8
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())
```


## Announcements

- Submit both your Rmd and the knitted HTML file. It helps us to review what you have done.
- Progress Check 1 coming next week
- A few words on peer evaluations


## Thinking about probabilities

Why learn probability?

- You could learn all the statistical analyses/tests without really understanding probability.


## Binomial probabilities

About 10% of people are left handed.

- What is the probability of exactly 0 left handed people in a *random sample* of 26?
- What is the probability of exactly 1?
- What is the probability of 1 *or more*?


## Binomial probabilities

Probability density functions in R: `dbinom()`, `dnorm()`, `dpois()`, etc.

`dbinom(x, size, prob)`

- `x` = Number of "successes"
- `size` = Sample size
- `prob` = Probability of success for each trial


## Binomial probabilities

What is the probability of *exactly* 0 left handed people in a *sample* of 26?

```{r eval=FALSE}
dbinom(0, 26, prob = 0.1)
```


## Binomial probabilities

What is the probability of *exactly* 0 left handed people in a *sample* of 26?

```{r}
dbinom(0, 26, prob = 0.1)
```


## Binomial probabilities

What is the probability of exactly 1?


## Binomial probabilities

What is the probability of exactly 1?

```{r}
dbinom(1, 26, prob = 0.1)
```


## Binomial probabilities

What is the probability of 1 *or more*?


## Binomial probabilities

What is the probability of 1 *or more*?

```{r}
sum(dbinom(1:26, 26, prob = 0.1))
```

*"or more"* is critical here for hypothesis testing:

- Probability of the observed data or any *more extreme*


## Binomial probabilities for Pr[left] = 0.1

```{r echo=FALSE}
M <- tibble(Lefties = 0:26,
            Probability = dbinom(0:26, 26, prob = 0.1))
ggplot(M, aes(Lefties, Probability)) + 
  geom_bar(stat = "identity") +
  labs(x = "n Left-handers") +
  scale_x_continuous(breaks = seq(0, 26, by = 5))
```

$n > 8$ have very low probabilities.


## Binomial probabilities for Pr[left] = 0.1

```{r echo=FALSE}
M <- tibble(Lefties = 0:26,
            Probability = dbinom(0:26, 26, prob = 0.1)) %>% 
  mutate(log_Prob = log10(Probability))
ggplot(M, aes(Lefties, log_Prob)) + 
  geom_bar(stat = "identity") +
  labs(x = "n Left-handers", y = "log10 Probability") +
  scale_x_continuous(breaks = seq(0, 26, by = 5))
```


## Powerball

1. 69 balls in one pool: Choose 5
1. 26 in the other pool: Choose 1

Number of winning combinations in pool 1 is the **binomial coefficient**

- the number of unique ordered sequences of successes that yield exactly $k$ successes from $n$ trials ("n choose k").
- the "successes" can happen in any order
- `choose(n, k)`

$$\left({n\atop k}\right)=\frac{n!}{k!\left(n - k\right)!}$$


## Powerball

1. 69 balls in one pool: Choose 5
1. 26 in the other pool: Choose 1

```{r}
choose(69, 5)
choose(26, 1)
```


## Powerball

5 successes from 69 chances $\times$ 1 success from 26 chances.

```{r}
choose(69, 5) * choose(26, 1)
```

What if the order had to be correct as well?


## Breast cancer screening

Approximately correct numbers:

1. 1 in 100 women in the age range 40-50 have breast cancer
2. 80% with cancer have a positive mammogram (true positives)
3. 10% without cancer have a positive mammogram (false positives)

What is the probability of a randomly sampled woman having breast cancer, given a positive test?


## Breast cancer screening

$$Pr[C|T] = \frac{Pr[T|C] \times Pr[C]}{Pr[T]}$$

- $Pr[T|C] = 0.8$
- $Pr[C] = 1/100 = 0.01$
- $Pr[T]$ = ???


## Breast cancer screening

$$Pr[C|T] = \frac{Pr[T|C] \times Pr[C]}{Pr[T]}$$

- $Pr[T|C] = 0.8$
- $Pr[C] = 1/100 = 0.01$
- $Pr[T] = (0.8 \times 0.01) + (0.1 \times 0.99)$


## Breast cancer screening

$$Pr[C|T] = \frac{Pr[T|C] \times Pr[C]}{Pr[T]}$$

- $Pr[T|C] = 0.8$
- $Pr[C] = 0.01$
- $Pr[T] = (0.8 \times 0.01) + (0.1 \times 0.99)$

```{r}
(0.8 * 0.01) / ((0.8 * 0.01) + (0.1 * 0.99))
```


## Breast cancer screening

Approximately correct numbers:

1. 1 in 100 women in the age range 40-50 have breast cancer
2. 20% with cancer have a negative mammogram (false negative)
3. 90% without cancer have a negative mammogram (true negative)

What is the probability of a randomly sampled woman having breast cancer, given a negative test?


## Breast cancer screening

$$Pr[C|N] = \frac{False~Negatives}{False~Negatives + True~Negatives}$$

$$Pr[C|N] = \frac{Pr[N|C] \times Pr[C]}{Pr[N]}$$

## Breast cancer screening

- 100 women
- 1 has cancer, 99 do not
- How many overall will get a negative test?

With cancer: 1 * 0.2 = 0.2


Without cancer: 99 * 0.9 = 89.1


Total negative tests: 0.2 + 89.1 = 89.3

Probability of having cancer, given negative test: 

```{r}
0.2/89.3
```
