---
title: "From Probability to Likelihood"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King and Kevin Middleton'
output:
  ioslides_presentation:
    fig_width: 8
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
library(forcats)
library(latex2exp)
```

## Frequency Distribution of Heads | 6 Flips

```{r echo=FALSE, fig.height=4.5}
flips <- tibble(
  Ways = c(1, 6, 15, 20, 15, 6, 1),
  Flips = c("0 H", "1 H", "2 H",
            "3 H", "4 H", 
            "5 H", "6 H")) %>% 
  mutate(Flips = fct_inorder(Flips))

flips %>% ggplot(aes(x = Flips, y = Ways)) +
  geom_bar(stat = "Identity")
```

## Probability Distribution of Heads | 6 Flips

```{r}
flips$Ways
flips$Ways / sum(flips$Ways)
```

## Probability Distribution of Heads | 6 Flips

```{r echo=FALSE, fig.height=3.5}
flip6 <- tibble(
  Ways = c(1, 6, 15, 20, 15, 6, 1),
  Flips = c("0 H", "1 H", "2 H",
            "3 H", "4 H", 
            "5 H", "6 H"),
  Prob = Ways / sum(Ways)) %>% 
  mutate(Flips = fct_inorder(Flips))

flip6 %>% ggplot(aes(x = Flips, y = Prob)) +
  geom_bar(stat = "Identity") +
  labs(y = "Probability")
```

How can we go from a set of probabilities to likelihoods?

## Model Likelihood ($\mathcal{L}$)

For a set of observations ($Y_i$) and hypothesized parameters ($\Theta$) the model likelihood is the product of the observations' individual likelihoods:

$$\mathcal{L}\left(\left\{ Y_{i}\right\} _{i=1}^{n};\Theta\right) = \prod_{i=1}^{n}\phi\left(Y_{i}; \Theta\right)$$

Evaluate the likelihood function for different values of $\Theta$ to estimate $\mathcal{L}$ for different sets of $\Theta$.

**Maximize $\mathcal{L}$ and you will have the maximum likelihood set of parameter estimates.**

## Model Likelihood ($\mathcal{L}$)

The product of a large number of probabilities can result in some very small numbers.

- Computers don't handle really small numbers very well
    - There is a lower limit to the smallest number a computer can keep track of.
    - This is $`r .Machine$double.xmin`$ in R.

Think about computing the model likelihood with thousands or millions of observations.

```{r}
.09^150
```

## Model Likelihood ($\mathcal{L}$)

Usually easier to maximize the (natural) log of the likelihoods.

- The log-likelihood is easier to deal with mathematically.

$$\mathcal{L}\left(\left\{ Y_{i}\right\} _{i=1}^{n};\Theta\right) = \prod_{i=1}^{n}\phi\left(Y_{i}; \Theta\right)$$

Log both sides of the equation:

$$\log\left(\mathcal{L}\left(\left\{ Y_{i}\right\} _{i=1}^{n};\Theta\right)\right) = \log\left(\prod_{i=1}^{n}\phi\left(Y_{i};\Theta\right)\right)$$

## Model Likelihood ($\mathcal{L}$)

Taking advantage of the algebraic rules associated with logs (the log of the products equals the sum of the logs):

$$\log\left(\mathcal{L}\left(\left\{ Y_{i}\right\} _{i=1}^{n};\Theta\right)\right) = \sum_{i=1}^{n} \log\left(\phi\left(Y_{i};\Theta\right)\right)$$

So we just need to sum the log-likelihoods of the observations to get the model likelihood. 

_Note_: `log()` is _natural_ log.

## Likelihood of coin flips

You have 3 heads from 6 coin flips (i.e., 1 observation)

- What is the maximum likelihood estimate of the probability of heads $(\theta)$?
- $\theta$ can be any value from 0 to 1 (but we won't include 0 or 1).

## log-Likelihood for one value of $\theta$

Hypothesize that $\theta = 0.2$, observe 3 heads from 6 flips. What is the likelihood?

```{r}
theta <- 0.2
pr <- dbinom(3, 6, prob = theta)
pr
log(pr)
```

## Likelihood across the range of 0 to 1

Generate a range of possible values for $\theta$:

```{r}
theta <- seq(0.001, 0.999, length = 200)
```

Calculate the probability for 3 heads for each value of $\theta$:

```{r}
pr <- dbinom(3, 6, prob = theta)
```

Convert to log-likelihoods:

```{r}
log_lik <- log(pr)
log_liks <- tibble(theta, log_lik)
```

---

```{r}
log_liks
```

## Likelihood across the range of 0 to 1

```{r echo=FALSE}
log_liks %>% 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```

## Maximum likelihood

```{r echo=FALSE}
log_liks %>% 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  geom_point(aes(x = theta[which.max(log_lik)], y = max(log_lik)),
             color = "red", size = 3) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```

## Maximum likelihood

```{r}
max(log_lik)
theta[which.max(log_lik)]
```

## Likelihood of coin flips

You have 20 heads from 60 coin flips.

- What is the maximum likelihood estimate of the probability of heads ($\theta$)?
- $\theta$ can be any value from 0 to 1.

## Likelihood across the range of 0 to 1

```{r}
theta <- seq(0.001, 0.999, length = 200)
pr <- dbinom(20, 60, prob = theta)
log_lik <- log(pr)
log_liks <- tibble(theta, log_lik)
```

## Likelihood across the range of 0 to 1

```{r echo=FALSE}
log_liks %>% 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```

## Maximum likelihood

```{r echo=FALSE}
log_liks %>% 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  geom_point(aes(x = theta[which.max(log_lik)], y = max(log_lik)),
             color = "red", size = 3) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```

## Maximum likelihood

```{r}
max(log_lik)
theta[which.max(log_lik)]
```

## Likelihood of coin flips

You have 20000 heads from 60000 coin flips.

- What is the maximum likelihood estimate of the probability of heads ($\theta$)?
- $\theta$ can be any value from 0 to 1.

## Likelihood across the range of 0 to 1

```{r}
theta <- seq(0.001, 0.999, length = 200)
pr <- dbinom(20000, 60000, prob = theta)
log_lik <- log(pr)
log_liks <- tibble(theta, log_lik)
```

## Maximum likelihood

```{r echo=FALSE}
log_liks %>% 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  geom_point(aes(x = theta[which.max(log_lik)], y = max(log_lik)),
             color = "red", size = 3) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```

## Quiz 04-2

Complete quiz 04-2.

Move on to Lecture 04-3.
