---
title: "Tail Probabilities"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King and Kevin Middleton'
date: 'Last updated: `r Sys.Date()`'
output:
  ioslides_presentation:
    fig_width: 8
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
source("Shade_Distributions.R")
```

## Notes


## Readings

## Recall from Problem Set 3

- Generate 100000 values from a normal distribution with a mean of 5 and a standard deviation of 1
- Generate 100000 values from a normal distribution with a mean of 5 and a standard deviation of 5

```{r}
set.seed(832476)
# 3
xx3 <- data_frame(num = rnorm(100000, 5, 1))

# 6
xx6 <- data_frame(num = rnorm(100000, 5, 5))
```

## Plots of random normal values

```{r echo=FALSE}
p1 <- xx3 %>% ggplot(aes(num)) +
  geom_histogram(bins = 50)
p2 <- xx6 %>% ggplot(aes(num)) +
  geom_histogram(bins = 50)

plot_grid(p1, p2, ncol = 2)
```


## What proportion of values from each set are above 7?

```{r}
#3 
nrow(xx3[xx3$num > 7, ]) / nrow(xx3)

#6
nrow(xx6[xx6$num > 7, ]) / nrow(xx6)
```

## Use `pnorm()` to calculate the probability of observing a value of 7 or greater for each distribution.

```{r}
pnorm(7, mean = 5, sd = 1, lower.tail = FALSE)
pnorm(7, mean = 5, sd = 5, lower.tail = FALSE)
```

Compare the values returned by `pnorm()` to those you calculated manually above.

## Tail probabilities

```{r}
shade_normal(0.025, tail = "both")
qt(0.025, 5)
```

## *t* Distribution

```{r}
shade_t(0.025, 5)
```

## *F* Distribution

```{r}
shade_F(0.05, 2, 10)
```

## $\chi^2$ Distribution

```{r}
shade_chisq(0.05, 2)
```

## Why do we always use 0.05 or 0.025?

5% is a special number  - the consensus threshold for statistical significance

<div class="columns-2">

```{r, echo=FALSE, fig.height=3.5,fig.width=3.5}
shade_normal(0.025, tail = "both")
```

```{r, echo=FALSE, fig.height=3.5,fig.width=3.5}
shade_normal(0.05, tail = "upper")
```

</center>

## We could choose any probability for any distribution

```{r, fig.height=3.5}
shade_normal(0.34, tail='upper')
```

## No Quiz

Watch Lecture 06-3

