---
title: "Confidence Intervals"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King and Kevin Middleton'
date: 'Last updated: `r Sys.Date()`'
output:
  ioslides_presentation:
    fig_width: 8
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
library(wesanderson)
library(rethinking)
source("Shade_Distributions.R")
```

## Confidence intervals measure the uncertainty in a parameter estimate

- Analytical  
    - Approximate calculation relying on 2 $\times$ SE 
    - Exact calculate using a critical *t*-value
- Maximum Likelihood 
    - Iterate over possible parameter values 
    - Identify the range of likelihoods summing to 0.95
- Bayesian Credible Intervals
    - Region including highest 95% of posterior density

## Calculating exact confidence intervals

<div class="columns-2">

```{r, echo=FALSE, fig.width = 4}
shade_normal(0.025)
```

- 2 is an approximation
- True value is ~1.96 for a normal distribution (infinite sample size/true population distribution)
- Samples rely on the *t*-distribution
- Increase width of CI for smaller sample size

</center>

## Calculating exact confidence intervals

```{r echo=FALSE}
shade_t(0.025, 5) +
  ggtitle(TeX("$t$-distribution with df = 5"))
```

Need the critical value that marks the 2.5% cutoff at either end.

## Calculating exact confidence intervals

Multiply the SE by the *t*-value encompassing 95% for a *t* distribution with a given # of degrees of freedom (e.g., a given sample size).

```{r t_plot, message=FALSE, fig.height=3.5, echo=FALSE, warning=FALSE}
x <- seq(-3, 3, by = 0.001)

sim <- data_frame(df = c(2, 10, 50)) %>%
  group_by(df) %>%
  do(data_frame(x = x, y = dt(x, .$df))) %>%
  mutate(Parameters = paste0("df = ", df)) %>%
  ungroup() %>%
  mutate(Parameters = factor(Parameters, levels = unique(Parameters)))

norm <- data_frame(
  x = x,
  y = dnorm(x, 0, 1)
)

pal <- wes_palette("Moonrise3", 3, type = "discrete")

ggplot() +
  geom_line(data = sim, aes(x, y, color = Parameters), size = 1.5) +
  geom_line(data = norm, aes(x, y)) +
  scale_color_manual(values = pal, name = "Degrees of\nFreedom") +
  labs(x = "x", y = "Relative Likelihood") +
  theme(legend.position = c(0.9, 0.75))
```

## Calculating exact confidence intervals

```{r}
qt(0.975, 10 - 1)

qt(0.975, 100 - 1)

qt(0.975, 1000000000 - 1)

qnorm(0.975, 0, 1)
```

## 95% CI for snake undulation (*n* = 8)

```{r}
undulation_rate <- c(0.9, 1.2, 1.2, 1.3, 1.4, 1.4, 1.6, 2.0)
undulation_mean <- mean(undulation_rate)
undulation_SEM <- sd(undulation_rate) / sqrt(length(undulation_rate))
(crit <- qt(0.025, df = 7, lower.tail = FALSE))
lower <- undulation_mean - crit * undulation_SEM
upper <- undulation_mean + crit * undulation_SEM
c(lower, undulation_mean, upper)
```

## Bayesian Credible Intervals

- Summarize the posterior distributions of Bayesian analyses
- Range of the most plausible estimates given the model
- XX% highest density interval
- XX% quantile

## MCMC sampling

- Most general way to do Bayesian inference
- No complicated integrals
- Requires time and computational power

## MCMC sampling

```{r echo=TRUE, cache=TRUE}
fm <- map2stan(
  alist(
    undulation_rate ~ dnorm(mu, sigma),
    mu ~ cauchy(0, 5),
    sigma ~ cauchy(0, 5)
  ),
  data = list(undulation_rate = undulation_rate),
  iter = 1e4,
  warmup = 1e3,
  WAIC = FALSE
)
```

## MCMC sampling output

```{r echo=FALSE, fig.height = 3}
rstan::traceplot(fm@stanfit, pars = c("mu", "sigma"))
```

## MCMC sampling

```{r echo=FALSE}
post <- as.data.frame(extract.samples(fm)) %>% 
  mutate(Sample = 1:length(mu))

post %>% 
  slice(1:20) %>% 
  gather(variable, value, -Sample) %>% 
  ggplot(aes(Sample, value)) +
  geom_line() +
  facet_grid(variable ~ ., scales = "free")
```

## MCMC method

- Current position (has a likelihood)
- Propose a change (has a likelihood)
- Move to proposed values with probability proportional to their ratio
- "Bad" jumps are possible with some non-zero probability

## Current estimates

Sample 1:

```{r}
post[1, ]
rel_liks <- dnorm(undulation_rate, mean = post$mu[1],
                  sd = post$sigma[1])
(current_ll <- sum(log(rel_liks)))
```

## Proposed estimates

```{r}
mu_prop <- 1.37
sigma_prop <- 0.3
rel_liks <- dnorm(undulation_rate, mean = mu_prop,
                  sd = sigma_prop)
(proposed_ll <- sum(log(rel_liks)))
```

```{r}
exp(proposed_ll) / exp(current_ll)
```

Ratio > 1, so definitely move.

```{r}
current_ll <- proposed_ll
```

## Proposed estimates

```{r}
rel_liks <- dnorm(undulation_rate, mean = 1.5, sd = 0.22)
proposed_ll <- sum(log(rel_liks))
exp(proposed_ll) / exp(current_ll)
```

Ratio < 1, so the probability of moving equal the ratio (here ~10%).

- This allows moves to lower likelihood estimates (as long as the probability of moving is not too small).

## MCMC sampling

The samples will oscillate around the most probable estimates, in proportion to their probabilities.

```{r echo=FALSE, fig.height = 4}
post <- as.data.frame(extract.samples(fm)) %>% 
  mutate(Sample = 1:length(mu))

post %>% 
  slice(1:100) %>% 
  gather(variable, value, -Sample) %>% 
  ggplot(aes(Sample, value)) +
  geom_line() +
  facet_grid(variable ~ ., scales = "free")
```

## Bayesian credible intervals

Posterior distribution of mean undulation, given, the model (including priors) and data:

```{r, echo=FALSE}
p <- post %>% 
  select(mu) %>% 
  ggplot(aes(mu)) +
  geom_line(stat = "density") +
  labs(y = "Relative Likelihood")
p
```

## Bayesian credible intervals

95% Highest density interval is the most probable location for the mean (given the model and the data).

```{r echo=FALSE, fig.height=3.5}
(cred <- HPDI(post$mu, prob = 0.95))
p + 
  geom_vline(xintercept = cred, color = "coral", size = 1.5)
```

## Interpreting Intervals

Frequentist interval: `r round(lower, 2)` -- `r round(upper, 2)`

- Future similar experiments and samples
- Does not indicate the probability of the parameter

Bayesian interval: `r round(cred[1], 2)` -- `r round(cred[2], 2)`

- These data and this model
- Relative probability within a region

## Quiz 06-3

Complete quiz 06-3.

Move on to Lecture 06-4.
