---
title: "Tail Probabilities"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King, Kevin Middleton, and Lauren Sullivan'
output:
  ioslides_presentation:
    fig_width: 8
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())
source("Shade_Distributions.R")
```

## Random normal values

- Generate 100,000 values from a normal distribution with a mean of 5 and a standard deviation of 1
- Generate 100,000 values from a normal distribution with a mean of 5 and a standard deviation of 5

```{r}
set.seed(832476)
sd1 <- tibble(y = rnorm(100000, 5, 1))
sd5 <- tibble(y = rnorm(100000, 5, 5))
```


## Plots of random normal values

```{r echo=FALSE}
p1 <- sd1 %>% ggplot(aes(y)) +
  geom_histogram(bins = 50) +
  labs(y = "Count")
p2 <- sd5 %>% ggplot(aes(y)) +
  geom_histogram(bins = 50) +
  labs(y = "Count")

plot_grid(p1, p2, ncol = 2)
```


## What proportion of values from each set are above 7?

```{r}
nrow(sd1[sd1$y > 7, ]) / nrow(sd1)
nrow(sd5[sd5$y > 7, ]) / nrow(sd5)
```


## Use `pnorm()` to calculate the probability of observing a value of 7 or greater for each distribution.

```{r}
pnorm(7, mean = 5, sd = 1, lower.tail = FALSE)
pnorm(7, mean = 5, sd = 5, lower.tail = FALSE)
```

Compare the values returned by `pnorm()` to those you calculated manually above.


## Tail probabilities

```{r fig.height = 3}
shade_normal(0.025, tail = "both")
qnorm(0.025)
```


## *t* Distribution

```{r fig.height = 3}
shade_t(0.025, df = 5)
qt(0.025, df = 5)
```


## *F* Distribution

```{r fig.height = 3}
shade_F(0.05, df1 = 2, df2 = 10)
qf(0.05, df1 = 2, df2 = 10, lower.tail = FALSE)
```


## $\chi^2$ Distribution

```{r fig.height = 3}
shade_chisq(0.05, df = 2)
qchisq(0.05, df = 2, lower.tail = FALSE)
```


## Why do we always seem to use 0.05 or 0.025?

5% is a special number - the arbitrary threshold for statistical significance

<div class="columns-2">

```{r, echo=FALSE, fig.height=3.5,fig.width=3.5}
shade_normal(0.025, tail = "both")
```

```{r, echo=FALSE, fig.height=3.5,fig.width=3.5}
shade_normal(0.05, tail = "upper")
```


## We could choose any probability for any distribution

```{r, fig.height=3.5}
shade_normal(0.34, tail='upper')
```

