---
title: "Unit 7: In Class Discussion"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King, Kevin Middleton, and Lauren Sullivan'
output:
  ioslides_presentation:
    fig_width: 8
    css: styles.css
---


```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())
source("https://raw.githubusercontent.com/kmiddleton/quant_methods/master/Lectures/Shade_Distributions.R")
```


## Quiz 7-4

- n = 25
- Mean: 98.524
- Standard deviation: 0.68

```{r}
n <- 25
y_bar <- 98.524
y_sd <- 0.68
```


## Standard Error of the Mean

$$SEM = \frac{\sigma}{\sqrt{n}}$$

```{r}
(SEM <- y_sd / sqrt(n))
```


## Test the sample mean vs. null hypothesized value of 98.6

$$t = \frac{\theta_1 - \theta_{null}}{SE_{\theta_1}}$$

```{r}
(t_stat <- (y_bar - 98.6) / SEM)
```


## Compare the t-statistic to the critical value ...

```{r}
(t_crit <- qt(p = 0.025, df = n - 1, lower.tail = TRUE))
t_stat
```

Is `t_stat` more extreme than `t_crit`?


## Compare the t-statistic to the critical value ...

```{r, fig.height=4}
shade_t(0.025, n - 1) +
  geom_vline(xintercept = t_stat, color = "goldenrod", linewidth = 1)
```


## 95% CI

```{r}
y_bar - qt(p = 0.025, df = n - 1, lower.tail = FALSE) * SEM
y_bar + qt(p = 0.025, df = n - 1, lower.tail = FALSE) * SEM
```


## How to figure out how many degrees of freedom?

How many parameters have you estimated?

- n - 1 (mean for one group)
- n - 2 (mean for 2 groups; slope and intercept)
- n - k (means for k groups)
