---
title: "Unit 10: In Class Discussion"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King, Kevin Middleton, and Lauren Sullivan'
output:
  ioslides_presentation:
    fig_width: 8
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
library(ggsci)
library(lme4)
library(AICcmodavg)

theme_set(theme_cowplot())
```

## What are nested models?

> For week 10 quiz 2, I have a question about the nested models. I was interpreting nested as everything stayed the same but you made it more complex but adding a variable or multiplying variables, so then I was confused when it went from dist_to_island in model 2, to area + dist_to_ecuador in model 3. Model 3 to 6 is what I was expecting (where a variable was added but none were changed, or addition turned into subtraction). Did I misinterpret the definition of nested?


## Nested models

```
fm0: N_Species ~ 1
fm1: N_Species ~ Dist_to_Island
fm2: N_Species ~ Area + Dist_to_Ecuador
fm3: N_Species ~ Dist_to_Island + Area + Dist_to_Ecuador
fm4: N_Species ~ Dist_to_Island + Elevation + Area + Dist_to_Ecuador
fm5: N_Species ~ Dist_to_Island + Elevation + Area * Dist_to_Ecuador
```


## Likelihood ratio test

```
Likelihood ratio test

Model 1: N_Species ~ 1
Model 2: N_Species ~ Dist_to_Island
Model 3: N_Species ~ Area + Dist_to_Ecuador
Model 4: N_Species ~ Dist_to_Island + Area + Dist_to_Ecuador
Model 5: N_Species ~ Dist_to_Island + Elevation + Area + Dist_to_Ecuador
Model 6: N_Species ~ Dist_to_Island + Elevation + Area * Dist_to_Ecuador

  #Df  LogLik Df   Chisq Pr(>Chisq)
1   2 -52.579                      
2   3 -52.450  1  0.2578     0.6116
3   4 -43.882  1 17.1357   3.48e-05
4   5 -43.857  1  0.0504     0.8224
5   6 -43.412  1  0.8903     0.3454
6   7 -43.404  1  0.0152     0.9020
```

## Akaike Weights

> I don't understand what the "weight" of each AIC model (the AICcWt column) actually tells us or how it's useful.

```
Model selection based on AICc:

    K   AICc Delta_AICc AICcWt Cum.Wt     LL
fm2 4 100.21       0.00   0.91   0.91 -43.88
fm3 5 105.21       5.01   0.07   0.99 -43.86
fm0 2 110.25      10.04   0.01   0.99 -52.58
fm4 6 110.82      10.61   0.00   1.00 -43.41
fm1 3 113.30      13.09   0.00   1.00 -52.45
fm5 7 119.48      19.27   0.00   1.00 -43.40
```


## Quiz 10-4


## Summary of model comparison procedures

1. Likelihood ratio tests
    - Significance tests of model differences
2. AIC and AICc
    - Model comparison that penalizes complexity (and small sample size)
3. Akaike Weights
    - Relative weighting of a set of models (0 - 100%)
4. Cross validation
    - Out-of-sample prediction error for a model


## Random effects model formulas

> What does the "~1| random" language mean?


## Salaries and experience example

```{r}
set.seed(238476287)

# Parameters for generating faculty salary data
departments <- c('Sociology', 'Biology', 'English', 'Informatics', 'Statistics')
base.salaries <- c(40000, 50000, 60000, 70000, 80000)
annual.raises <- c(2000, 500, 500, 1700, 500)
faculty.per.dept <- 20
total.faculty <- faculty.per.dept * length(departments)

# Generate dataframe of faculty and (random) years of experience
ids <- 1:total.faculty
Department <- rep(departments, faculty.per.dept)
Experience <- floor(runif(total.faculty, 0, 10))
Bases <- rep(base.salaries, faculty.per.dept) * runif(total.faculty, .9, 1.1) # noise
Raises <- rep(annual.raises, faculty.per.dept) * runif(total.faculty, .9, 1.1) # noise
M <- data.frame(ids, Department, Bases, Experience, Raises)

# Generate salaries (base + experience * raise)
M <- M %>% mutate(
    Salary = Bases + Experience * Raises,
    Department = fct_relevel(Department, "Informatics")) %>% 
  select(-ids, -Bases, -Raises)
```


## Visualization

```{r, echo=FALSE}
p1 <- ggplot() +
  # geom_vline(xintercept = 4.5, color = "gray", linetype = "dashed") +
  # geom_hline(yintercept = mean(M$Salary), color = "gray", linetype = "dashed") +
  geom_point(data = M, aes(x = Experience, y = Salary, color = Department),
             size = 3, alpha = 0.5) +
  scale_color_jama() +
  labs(x = "Years of Experience", y = "Salary")
p1
```


## Means by department

```{r}
M %>% 
  group_by(Department) %>% 
  summarise(Salary = mean(Salary),
            .groups = "drop")
```

## Model with constant salary

```{r}
m0 <- lm(Salary ~ -1 + Department, data = M)
coef(m0)
```

## Predictions

```{r}
Pred <- crossing(Experience = 0:9,
                 Department = levels(M$Department)) %>% 
  mutate(m0 = predict(m0, newdata = .))
Pred %>% arrange(Department, Experience) %>% slice(1:10)
```

## Model with constant salary

```{r, echo=FALSE}
p2 <- p1 +
  geom_line(data = Pred, aes(x = Experience, y = m0, color = Department),
            linetype = "dotted")
p2
```


## Model without respect to grouping

```{r}
m1 <- lm(Salary ~ Experience, data = M)
coef(m1)
Pred$m1 <- predict(m1, newdata = Pred)
Pred %>% arrange(Experience) %>% slice(1:10)
```


## Model without respect to grouping

```{r, echo=FALSE}
p2 +
  geom_line(data = Pred, aes(x = Experience, y = m1),
            size = 2)
```


## Additive multiple regression model

```{r}
m2 <- lm(Salary ~ Experience + Department, data = M)
coef(m2)
Pred$m2 <- predict(m2, newdata = Pred)
Pred %>% arrange(Experience) %>% slice(1:10)
```


## Additive multiple regression model

```{r, echo=FALSE}
p2 +
  geom_line(data = Pred, aes(x = Experience, y = m2, color = Department),
            size = 2)
```


## Multilevel model with varying intercept

```{r}
m3 <- lmer(Salary ~ Experience + (1 | Department), data = M, REML = FALSE)
Pred$m3 <- predict(m3, newdata = Pred)
Pred %>% arrange(Experience) %>% slice(1:10)
```


## Multilevel model with varying intercept

```{r, echo=FALSE}
p2 +
  geom_line(data = Pred, aes(x = Experience, y = m3, color = Department),
            size = 2)
```


## Compare varying intercept models

```{r, echo=FALSE}
p2 +
  geom_line(data = Pred, aes(x = Experience, y = m2, color = Department),
            size = 0.5, linetype = "dashed") +
  geom_line(data = Pred, aes(x = Experience, y = m3, color = Department),
            size = 0.5)
```


## Multilevel model with varying slope

```{r}
m4 <- lmer(Salary ~ Experience + (0 + Experience | Department), data = M,
           REML = FALSE)
Pred$m4 <- predict(m4, newdata = Pred)
```


## Multilevel model with varying slope


```{r, echo=FALSE}
p2 +
  geom_hline(yintercept = mean(M$Salary)) +
  geom_line(data = Pred, aes(x = Experience, y = m4, color = Department),
            size = 2)
```


## Interaction model

```{r}
m5 <- lm(Salary ~ Experience * Department, data = M)
Pred$m5 <- predict(m5, newdata = Pred)
```


## Interaction model

```{r, echo=FALSE}
p2 +
  geom_line(data = Pred, aes(x = Experience, y = m5, color = Department),
            size = 2)
```


## Multilevel model with varying slope and intercept

```{r}
m6 <- lmer(Salary ~ Experience + (1 + Experience | Department), data = M,
           REML = FALSE)
Pred$m6 <- predict(m6, newdata = Pred)
```


## Multilevel model with varying slope and intercept

```{r, echo=FALSE}
p2 +
  geom_line(data = Pred, aes(x = Experience, y = m6, color = Department),
            size = 2)
```


## Compare varying slope and intercept models

```{r, echo=FALSE}
p2 +
  geom_line(data = Pred, aes(x = Experience, y = m5, color = Department),
            size = 0.5, linetype = "dashed") +
  geom_line(data = Pred, aes(x = Experience, y = m6, color = Department),
            size = 0.5)
```


## Model comparison: `lm()`

```{r, message=FALSE}
library(AICcmodavg)
aictab(list(m0, m1, m2, m5), modnames = c("m0", "m1", "m2", "m5"))
```


## Model comparison: `lmer()`

```{r}
aictab(list(m3, m4, m6), modnames = c("m3", "m4", "m6"))
```

