---
title: "Distribution Free Methods: Introduction"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King and Kevin Middleton'
output:
  ioslides_presentation:
    fig_width: 8
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
library(readxl)
library(wesanderson)
```

## Readings

North BV, Curtis D, Sham PC. [-@North2002-nu]. A note on the calculation of empirical P values from Monte Carlo procedures. *Am J Hum Genet* 71:439-441.

## Additional Resources

- Ludbrook J, Dudley H. [-@Ludbrook1998-zg]. Why Permutation Tests Are Superior to t and F Tests in Biomedical Research. Am Stat 52:127–132.
- Manly BFJ. [-@Manly2006-ky]. *Randomization, Bootstrap and Monte Carlo Methods in Biology*. 3rd edition. Boca Raton, FL: Chapman and Hall/CRC.
- Roff DA. [-@Roff2006-lz]. *Introduction to Computer-intensive Methods of Data Analysis in Biology*. Cambridge University Press.

## Distribution Free Methods

Distribution free methods do not assume the data are drawn from any specific probability distribution. *Important: distribution-free, not assumption-free (e.g., IID).*

  - Randomization 
      - Shuffle data set in some way
    
  - Resampling
      - Sample from data set in some way
        - Jackknife
        - Bootstrap

## Why distribution free methods?      
  
  - Real data often does not follow standard distributions
  - Very flexible
  - Uses only information from your data set
      - Good & Bad
  - Standard errors can not always be calculated for all parameters
  
## Resampling

  - Primarily a method for parameter estimation and interval estimation
  - Sample from the data to estimate the variability in the parameter estimate
  - **Assumption:** Observed distribution is representative of the true distribution
  - Need to empirically show the method works for each type of application using simulation

## Resampling procedure: bootstrap

  > Bootstrap: Sample N observations with replacement, estimate parameter, repeat k times

```{r echo=FALSE, fig.width=7, fig.height=4}
set.seed(519)

BeakLength <- data.frame("BeakLength" = rnorm(100, 12, 0.5))

ggplot(BeakLength, aes(x = BeakLength)) +
  geom_histogram(bins = 20) +
  geom_vline(xintercept = mean(BeakLength$BeakLength),
             color = 'coral', lwd = 3)
```

## Resampling procedure: bootstrap

- Sample with replacement, n = number of observations

```{r echo=FALSE, fig.width=7, fig.height=4}
bs1 <- sample(BeakLength$BeakLength, replace = TRUE)
cat(bs1[1:6], "...")
```

- Get mean of sample (& repeat):

```{r echo=FALSE, fig.width=5, fig.height=3, fig.align='center'}
mean(bs1)

BS <- data.frame('BS' = numeric(length = 100))

for (jj in 1:100) {
  BS[jj,'BS'] <- mean(sample(BeakLength$BeakLength, replace = TRUE))
}

ggplot(BS, aes(x = BS)) +
  geom_histogram(bins = 20)
```
 
## Resampling procedure: bootstrap 

- Estimate of mean = mean of the bootstrap estimates
- Estimate of SE = standard deviation of the bootstrap estimates

```{r echo = FALSE, fig.width=7, fig.height=4}
ggplot(BS, aes(x = BS)) +
  geom_histogram(bins = 20) +
  geom_vline(xintercept = mean(BS$BS), color = 'coral', lwd = 3) +
  geom_vline(xintercept = (mean(BS$BS) - sd(BS$BS)), color = 'slateblue') +
  geom_vline(xintercept = (mean(BS$BS) + sd(BS$BS)), color = 'slateblue')
```

## Resampling procedure: jackknife

> Jackknife: Remove 1 observation, estimate parameter, repeat for each observation in the data set, removing each exactly once.

```{r echo=FALSE, fig.width=7, fig.height=4}
ggplot(BeakLength, aes(x = BeakLength)) +
  geom_histogram(bins = 20) +
  geom_vline(xintercept = mean(BeakLength$BeakLength),
             color = 'coral', lwd = 3)
```

## Resampling procedure: jackknife

```{r echo=FALSE}
cat("Obs:",BeakLength$BeakLength[1:6], "...")
cat(" \t\tX\t",BeakLength$BeakLength[2:6], "...")
cat("  \t",BeakLength$BeakLength[1],"\tX\t ", BeakLength$BeakLength[3:6], "...")
cat("....")
```

## Applications & Resources: Resampling

1. Bootstrapping in phylogenetics: Sample character vectors, construct best tree, repeat. [@Felsenstein1985-my; @Felsenstein1993-on]

1. Jackknifing in quantitative genetics [@Roff1994-gz; @Roff2002-rf; @Roff2008-on]

## Randomization

  - Primarily a method for hypothesis testing
  - More directly asks: What is the probability of observing a result as extreme or more than my observed result? 
  
  **Assumption:** Under the null hypothesis, observations are random draws from a common population

## Randomization procedure

1. Decide on a test statistic
1. Calculate the test statistic for the *observed* data
1. Randomly shuffle the observations
1. Calculate the test statistic for that group
1. Repeat thousands of times
1. Determine the proportion of random combinations resulting in a test statistic more extreme than the observed value ("empirical *P*")

When test assumptions are met, the results will match asymptotic procedures. When the assumptions are not met, the results will be valid.

## Decide on a test statistic

1. Mean difference
2. *t*: *t*-test, linear model parameter estimate (slope, intercept)
3. *F*: ANOVA-like
4. $\chi^2$
5. Any metric of your choice (P-value, Fst, heterozygosity, LOD score, etc.)


## Randomization procedure

```{r echo=FALSE, fig.height=4}
set.seed(903)
dat <- data.frame('BeakLength' = c(rnorm(20, 12, 0.5),
                                   rnorm(20, 13.5, 0.5)),
                  "Original" = rep(c('A', 'B'), each = 20),
                  "Type" = rep(c('A', 'B'), each = 20))

cat("Mean A:", mean(dat$BeakLength[dat$Type == 'A']),
    "\nMean B:", mean(dat$BeakLength[dat$Type == 'B']))

ggplot(dat, aes(x = Type, y = BeakLength, color = Original)) +
  geom_point(position = position_jitter(width = 0.05),
             alpha = 1 / 2, size = 2)
```

## Randomization procedure

Shuffle Type

```{r echo=FALSE, fig.height=4}
dat.s <- data.frame('BeakLength' = dat$BeakLength,
                    "Original" = dat$Original,
                    "Type" = sample(dat$Type))

cat("Mean A:", mean(dat.s$BeakLength[dat.s$Type == 'A']),
    "\nMean B:", mean(dat.s$BeakLength[dat.s$Type == 'B']))

ggplot(dat.s, aes(x = Type, y = BeakLength, color = Original)) +
  geom_point(position = position_jitter(width = 0.05),
             alpha = 1 / 2, size = 2)
```

## Randomization procedure

Repeat N times

```{r echo=FALSE, fig.height=4}
dat.s <- data.frame('BeakLength' = dat$BeakLength,
                    "Original" = dat$Original,
                    "Type" = sample(dat$Type))

cat("Mean A:", mean(dat.s$BeakLength[dat.s$Type == 'A']),
    "\nMean B:", mean(dat.s$BeakLength[dat.s$Type == 'B']))

ggplot(dat.s, aes(x = Type, y = BeakLength, color = Original)) +
  geom_point(position = position_jitter(width = 0.05),
             alpha = 1 / 2, size = 2)
```

## Quiz 10-1

Lecture 10-2

## References
