---
title: "Experimental Design: Power II"
author: 'Special Topics: Multivariate Statistics'
date: 'Last updated: `r Sys.Date()`'
output:
  ioslides_presentation:
    fig_width: 7
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
library(readxl)
knitr::opts_chunk$set(cache = TRUE)
```

## Power analysis via simulation

```{r}
library(pwr)
pwr.t.test(power = 0.8, d = 0.6)
```

## Power analysis via simulation

Estimate power across a range of *d* and *n*:

`d <- seq(0.1, 2, length.out = 50)`

- Standardized mean differences from 0.1 to 2

`n <- c(5, 10, 15, 20, 30, 50, 100)`

- Sample sizes from 5 to 100

## Power analysis via simulation

```{r pwr_sim}
d_seq <- seq(0.1, 2, length.out = 50)
ns <- c(5, 10, 15, 20, 30, 50, 100)

# data.frame with all combinations on d_seq and ns
power_t <- expand.grid(d_seq, ns)
names(power_t) <- c("d", "n")

# Column for calculated power
power_t$power <- numeric(nrow(power_t))

for (i in 1:nrow(power_t)) {
  d <- power_t$d[i]
  n <- power_t$n[i]
  power_t$power[i] <- pwr.t.test(n, d)$power
}
```

## Power analysis via simulation

```{r echo=FALSE}
ggplot(power_t, aes(d, power, color = factor(n))) +
  geom_hline(yintercept = 0.8, color = "darkblue") +
  geom_line(size = 2) +
  geom_point(aes(x = 0.6, y = 0.8), color = "red") +
  ylim(c(0, 1)) +
  scale_color_discrete(name = "n") +
  theme(legend.justification = c(1, 0), legend.position = c(1, 0))
```

## Power analysis via simulation

Generalizing to any type of analysis:

1. Specify parameters (e.g., *n*, *d*, means, slope, intercept) in a biologically realistic range (reasonable variances)
1. Simulate data that your process produces
1. Determine if *P* value is less than $\alpha$.
1. Repeat

## Power analysis via simulation

```{r pwr_manual_sim}
set.seed(20)

nsims <- 1e4
d <- 0.4
n <- 15
alpha <- 0.05
sig <- logical(nsims)  # Vector of TRUE/FALSE

for (i in 1:nsims) {
  x1 <- rnorm(n, mean = 0, sd = 1)
  x2 <- rnorm(n, mean = d, sd = 1)
  p <- t.test(x1, x2, var.equal = TRUE)$p.value
  sig[i] <- p < alpha
}
```

## Power analysis via simulation

```{r}
head(sig)
mean(sig)
pwr.t.test(n = 15, d = 0.4)$power
```

## Power analysis via simulation

Simulate data from a exponential process:

$$\mbox{Femur Length} = a \mbox{Mass}^b$$

$$\log \mbox{Femur Length} = \log a + b \log \mbox{Mass}$$

What is the power to detect deviations from isometry?

- Simulate across a range of *n*
- Use a range of slopes from 1/3 - 0.2 to 1/3 + 0.2

## Power analysis via simulation

```{r}
nsims <- 1e4
alpha <- 0.05
ns <- c(5, 10, 25, 50, 100, 200, 400)
b_null <- 1/3
b_devs <- seq(-0.2, 0.2, length.out = 100)

# Create data.frame with all combinations of ns and b_devs
pwr_reg <- expand.grid(ns, b_devs)
names(pwr_reg) <- c("n", "b_dev")
pwr_reg$Power <- NA
```

## Power analysis via simulation  {.smaller}

```{r pwr_sma, eval=FALSE}
set.seed(912)
# Iterate through the rows of `pwr_reg`
for (i in 1:nrow(pwr_reg)) {
  tic <- Sys.time()
  message(i, " of ", nrow(pwr_reg))
  n <- pwr_reg$n[i]
  b_dev <- pwr_reg$b_dev[i]
  sig <- logical(nsims)
  
  for (j in 1:nsims) {
    log_Mass <- log(runif(n, 1, 1e3))
    log_a <- rnorm(n, 1.31, 0.15)
    log_Fem_Len <- log_a + (b_null + b_dev) * log_Mass
    fm <- sma(log_Fem_Len ~ log_Mass, slope.test = b_null, method = "OLS")
    sig[j] <- fm$slopetest[[1]]$p < alpha
  }
  pwr_reg$Power[i] <- mean(sig)
  save(pwr_reg, file = "./data/pwr_reg_SMA.Rda")
  message(Sys.time() - tic)
}
```

- Mass uniformly distributed from 1 - 1000
- $a$ normally distributed with a mean of 1.31
- Calculate `log_Fem_Len`

## Power analysis via simulation

- 7,000,000 regressions
- ~8 hours later...

## Power analysis via simulation

```{r echo=FALSE}
load("../data/pwr_reg_SMA.Rda")
p <- ggplot(pwr_reg, aes(b_dev, Power, color = as.factor(n))) +
  geom_line() +
  geom_line(size = 2) +
  ylim(c(0, 1)) +
  geom_hline(yintercept = 0.8, color = "blue", linetype = "dotted") +
  geom_hline(yintercept = 0.05, color = "red", linetype = "dotted") +
  scale_color_discrete(name = "n") +
  labs(x = "Slope Deviation") +
  theme(legend.justification = c(1, 0), legend.position = c(1, 0))
p + theme(legend.background = element_rect(fill = "gray85", 
    linetype = "solid"))
```

## Quiz 14-3

End of lectures.
