---
title: "Progress Check 1"
subtitle: "Quantitative Methods in Life Sciences"
author: 'Elizabeth King, Kevin Middleton and Lauren Sullivan'
output:
  ioslides_presentation:
    fig_width: 7
    css: styles.css
csl: evolution.csl
bibliography: Multivariate.bib
---

```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())
```


## `runif()` vs. `dunif()`

> Could you explain the difference between runif and dunif? When generating values for PS_04 they do not return the same kind of distribution even though it seems like they should both return uniform distributions of values.

## Probability mass

```{r, echo=FALSE}
tibble(y = 0:6,
       d = dbinom(y, size = 6, prob = 0.5)) %>% 
  ggplot(aes(y, d)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = 0:6)+
  labs(title = "Binomial(0:6, 6, 0.5)",
       y = "Density",
       x = "Successes")
```


## Probability density

```{r, echo=FALSE}
tibble(y = seq(-5, 5, length.out = 200),
       d = dnorm(y)) %>% 
  ggplot(aes(y, d)) +
  geom_line() +
  labs(title = "Standard Normal Distribution",
       y = "Density")
```


## `runif()` vs. `dunif()`

```{r}
set.seed(2389749)

runif(n = 10, min = 0, max = 1)
```

1,000,000 random numbers [0, 1]

```{r}
y <- tibble(y = runif(n = 1e6, min = 0, max = 1))
```


## `runif()` vs. `dunif()`

```{r, echo=FALSE}
ggplot(y, aes(y)) +
  geom_histogram(bins = 20)
```


## `runif()` vs. `dunif()`

```{r, echo=FALSE}
ggplot(y, aes(y)) +
  geom_density() +
  labs(y = "Density")
```


## `runif()` vs. `dunif()`

```{r}
d <- tibble(
  y = seq(0, 1, length.out = 200),
  d = dunif(y, min = 0, max = 1))
head(d)
```


## `runif()` vs. `dunif()`

```{r, echo=FALSE}
ggplot(d, aes(y, d)) +
  geom_line() +
  labs(title = "min = 0, max = 1",
       y = "Density")
```


## `runif()` vs. `dunif()`

```{r}
d <- tibble(
  y = seq(0, 10, length.out = 200),
  d = dunif(y, min = 0, max = 10))
```

What is the density in the range $0 \rightarrow 10$?


## `runif()` vs. `dunif()`

```{r, echo=FALSE}
ggplot(d, aes(y, d)) +
  geom_line()  +
  labs(title = "min = 0, max = 10",
       y = "Density")
```
