---
title: "Untitled"
author: "KingMiddleton"
date: "4/5/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nlme)
library(cowplot)
```

## Spatial Non-independence

At the Conard Environmental Research Area, there are 20 experimental plots with 10 burned each fall and 10 unburned. The following data are grass species richness data from 5 separate, hapazardly chosen, 10 meter^2  plots within each of these 20 experimental plots. The hypothesis to be tested is that burning increases species richness.


```{r }
#set.seed(54221)
set.seed(112)
mms<-c(rnorm(10, 5.5, 2),rnorm(10,6.5,2))
all.rich<-numeric(0)
for(ii in mms){
all.rich<-c(all.rich,as.integer(rnorm(5,ii,1)))
}

rich.dat<-data.frame('plot'=as.factor(rep(seq(1,20),each=5)), 
                     'treatment'= rep(c('burned','unburned'),each=50), 
                     'richness'=all.rich)
```

### Activity

Load in the richness data (NAME). Plot species richness vs. treatment. Include an appropriate amount of jitter and transparancy and color points by plot. 

```{r}

ggplot(rich.dat, aes(x=treatment, y=richness, color=plot)) +
  geom_point(position = position_jitter(0.1), alpha=1/2)


```



Do you think there is a relationship between species richness and burning?

>

First, fit a model predicting species richness from treatment without considering plot membership. What can you conclude from the results of this model?

>

```{r}

anova(lm(richness~treatment, data=rich.dat))

```

Now, get the mean species richness value for each plot. Fit a model predicting mean species richness from treatment. What can you conclude from this model?

```{r}

mean.plot<-rich.dat %>% group_by(plot,treatment) %>% summarize(meanrich=mean(richness))
anova(lm(meanrich~treatment, data=mean.plot))


```

>

Use the `lme()` function to fit a model predicting richness from treatment with plot as a random effect. 

```{r}

summary(lme(richness ~ treatment, random=~1|plot, data=rich.dat))

```

Compare your results to the approach using the mean value. What can you conclude from this model?


These 20 plots exist in an alternating pattern going from south to north on either side of an access road. What other source of non-independence should you consider? How would you account for this factor?

> 



