---
title: 'Problem Set 04'
author: "Your Name Here: Group X"
date: 'Last updated: `r Sys.Date()`'
output:
  html_document:
    smart: no
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(cowplot)
```

## Flow control: for loops 1

For loops are one of the main methods for iterating through R objects (vectors, data.frames, etc.).^[do ... while constructions are much less common]. Study and run the following code:

```{r}
for (ii in 1:10) {
  cat(ii, "\n")
}
```

There are a few things to look at here:

- The `for ()` part of the statement set up what will be iterated through (the sequence 1 to 10). At each iteration `ii` takes the next value and can be used within in the loop.
- `ii` is a better choice of iterator name than `i`, because it is easier to search for later. Imagine trying to find `i` in your code.
- `cat()` is a function ("concatenate") that you can use to print output to the screen. The statement here tells R to print each value of `ii` followed by a new line (`"\n"`).
- Until you get comfortable with for loops, it's often useful to print out the values of the iterator and/or other critical data with each iteration so that you are certain of their values.

### Activity

Write you own for loop, choosing a new variable name for the iterator, which will start at 50 and end at 60. Within the loop, print the value of the iterator and add 10.

```{r}

```

Your output should be the sequence 60 to 70.

## Flow control: for loops 2

A common way to work with for loops is to iterate through a pre-existing R object. The for loop will take on the values starting with the first value and iterating through to the last value of the object. So far, we have used sequence by 1, but you can iterate through the values of a vector. Here you will just pass the vector name: `for (ii in myvector)`.

### Activity

Create a vector (`myvector`) that contains the sequence of the odd numbers 1 through 21.

```{r}

```

Iterate through `myvector`, printing out the value of the iterator and the square of the iterator. Separate the values inside `cat()` with a tab (`"\t"`).

```{r}

```

Repeat the same loop, but iterate from 1 to the `length()` of `myvector`. Inside the loop, extract the element of `myvector` that corresponds to the value of the iterator. This construction, where the first line inside the loop extracts a value from the iterator, is much more common, because more often than not, you will need the value of the iterator later in the loop.

```{r}

```

Notice here that the value of the iterator goes from 1 to 11 (the length of `myvector`).

## Functions

Functions are reusable pieces of code that perform operations. Although functions can be very complex, best coding practice is to keep functions short and simple. In R, you create functions with the function `function()`.

Below is a function to add 10 to whatever object is passed to it.

```{r}
add_10 <- function(x) {
  y <- x + 10
  return(y)
}
```

There are a few things to note here:

- Anything within the `()` after `function` are arguments that are passed to the function. In this function, we require an argument `x`. There are mechanisms for allowing optional arguments and specifying default values. You can read about them one (see the R programming resources on Canvas).
- The variable `y` exists inside the function only. It does not move outside the function, until you explicitly say so: `return(y)`.
- We haven't done any kind of checking to see whether the argument `x` can be added to. If we pass a character vector as `x`, we will get an error. `non-numeric argument to binary operator`. Down the road, you might consider checking the incoming argument(s).

```{r}
zz <- 100
add_10(x = zz)
add_10(zz)

add_10(seq(50, 60, by = 2))

# y doesn't change due to the function
y <- 5412
add_10(y)
y

# y is reassigned to the result of the function
y <- add_10(y)
y
```

### Activity

Write your own function below that takes two arguments. Multiply the arguments and return the value. Try out your function to make sure that it works for both single numerics and numeric vectors.

```{r}

```

## Sometimes I think George is not being completely honest with me

You and your friend George go to dinner once a week. You agreed that you would flip George's "lucky" coin to decide who gets to choose the restaurant. Being superstitious, you always pick heads

After losing several times in a row, you start to get suspicious and begin keeping track. You find that after 20 weeks, you have only chosen the restaurant 5 times.

### Activity

What is the maximum likelihood estimate (MLE) of the probability of heads for this sample from this coin? Test values ($\theta$) from 0.05 to 0.95 in increments of 0.05, and plot the log-likelihood vs. the values for $\theta$.

```{r}

```

Use `which.max()` to extract the value of $\theta$ that maximizes the log-likelihood. Refer to the lecture slides if you need a reminder.

```{r}

```

Re-plot the data and add a large, colorful point with `geom_point()` located at the MLE. The full list of colors available in R is at http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf 

```{r}

```

Under certain circumstances the ratio of two probabilities (likelihoods) can be interpreted as a test of one model compared to another. In this case we have two models, (1) the maximum likelihood estimate of the probability of heads and (2) the null model that the coin is fair ($Pr[Heads] = 0.5$). 

Calculate the probabilities (not log-likelihoods) of observing 5 heads given:

1. Your maximum likelihood estimate above. 
2. The null probability of 0.5

```{r}

```

Now calculate the ratio of the first model to the second. This quantity is called a *Bayes Factor*, which can be interpreted as the strength of the evidence of one model compared to another.

```{r}

```

Interpretations of Bayes Factors are:^[Kass R, Raftery AE. 1995. Bayes factors. *J Am Stat Assoc* 90:773-795.] 

1. $< 3.2$: Not worth a mention
2. $3.2 - 10$: Substantial
3. $10 - 100$: Strong
4. $> 100$: Decisive

What is you interpretation of the Bayes Factor for the MLE vs. the null probability of 0.5?

> 

What is the probability of observing these data or anything more extreme (fewer heads)?

```{r}

```

You win the next two coin flips and get to choose the restaurant. Update your calculations of the Bayes Factor and probability as above.

```{r}

```

What do you conclude about the coin now?

> 
