---
title: 'Problem Set 14'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
output:
  html_document:
    smart: no
    theme: flatly
---

```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
# FIXME
library(readxl)
library(tidyverse)
library(cowplot)
library(pwr)
```

## Planning experiments

Consider that you are planning some experiments, use the `pwr` package to calculate the unknown quantity for each of the following situations. Assume that $\alpha$ = 0.05 for all tests.

### Activity

Use `cohen.ES()` to look up the effect size for a "small" effect for a *t*-test.

```{r}
# FIXME
cohen.ES(test = "t", size = "small")
```

Calculate the sample size (*n*) needed in each group of a two-sample *t*-test with power = 0.80 to detect a small effect (use the effect size from above).

```{r}
# FIXME
pwr.t.test(d = 0.2, power = 0.80, type = "two.sample")
```

Repeat the test above but for a paired *t*-test.

```{r}
# FIXME
pwr.t.test(d = 0.2, power = 0.80, type = "paired")
```

Calculate the number of observations for a correlation test where you estimate a correlation coefficient of 0.6. Power should be 0.80.

```{r}
# FIXME
pwr.r.test(r = 0.6, power = 0.8)
```

Calculate the power for a correlation test where you estimate the correlation coefficient to be 0.4, for a sample size of 15.

```{r}
# FIXME
pwr.r.test(r = 0.4, n = 15)
```

## Calcium concentration in bird plasma

The file `Bird_Plasma.xlsx` contains factorial data on blood plasma calcium concentration (`Calcium`, mg / 100 mL) in male and female birds (`Sex`) each of which was treated or not with a hormone (`Hormone`).

### Activity

Load the Excel file, and convert treatment and sex to factors.

```{r}
# FIXME
M <- read_excel("../data/Bird_Plasma.xlsx")
M <- M %>% mutate(Treatment = factor(Treatment),
                  Sex = factor(Sex))
```

Use `group_by()` and `tally()` to determine the sample size in each of the groups.

```{r}
# FIXME
M %>% group_by(Treatment, Sex) %>% tally()
```

Assume that the researchers who carried out this study had performed a power analysis prior to collecting data. Calculate the minimum effect size they could detect with their experimental design. Assume that there are 4 groups and power is 0.80.

```{r}
# FIXME
pwr.anova.test(k = 4, n = 5, power = 0.80)
```

How do you interpret the minimum effect size that the researchers could detect at power of 0.8?

> The effect has to be really large (> 0.8 standard deviations) in order to detect it.

## 
